---
title: "class14"
author: "Joshua Martin (PID:A18545389)"
format: pdf
toc: TRUE
---

## Section 1. Differential Expression Analysis
```{r}
library(DESeq2)

```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)

# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

```{r}
countData <- as.matrix(countData[ , -1])
head(countData)
```
>Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

rowSums returns the total read count for each gene across the samples. rowSums = 0 means that gene wasn't detected in any of the samples.

```{r}
countData <- countData[rowSums(countData) > 0, ]
head(countData)
```

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```
```{r}
dds

```
> Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
summary(res)
```
> Q. Improve this plot by completing the below code, which adds color and axis labels


```{r}
mycols <- rep("gray", nrow(res))

mycols[abs(res$log2FoldChange) > 2] <- "red"

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[inds] <- "blue"

plot(res$log2FoldChange,
     -log(res$padj),
     col = mycols,
     xlab = "Log2(FoldChange)",
     ylab = "-Log(P-value)")
```

> Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.



```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol <- mapIds(org.Hs.eg.db,
                     keys    = row.names(res),
                     keytype = "ENSEMBL",
                     column  = "SYMBOL",
                     multiVals = "first")

res$entrez <- mapIds(org.Hs.eg.db,
                     keys    = row.names(res),
                     keytype = "ENSEMBL",
                     column  = "ENTREZID",
                     multiVals = "first")

res$name <- mapIds(org.Hs.eg.db,
                   keys    = row.names(res),
                   keytype = "ENSEMBL",
                   column  = "GENENAME",
                   multiVals = "first")

head(res, 10)
```

>Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory. 

```{r}
res <- res[order(res$pvalue), ]
write.csv(res, file = "deseq_results.csv")
```


## Section 2: Pathway Analysis

```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)

attributes(keggres)
```

```{r}
head(keggres$less)
```

```{r}
library(pathview)
pathview(gene.data=foldchanges, pathway.id="hsa04110")

```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)

```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

> Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?

```{r}

keggrespathways_down <- rownames(keggres$less)[1:5]


keggresids_down <- substr(keggrespathways_down, start = 1, stop = 8)
keggresids_down
```

```{r}
pathview(gene.data = foldchanges,
         pathway.id = keggresids_down,
         species   = "hsa")
```

## Section 3. Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

## Section 4. Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

>Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Cell Cycle (p = 2.65 x 10^-5), followed by
Cell Cycle, Mitotic
Cell Cycle Checkpoints
Mitotic Prometaphase
Mitotic Spindle Checkpoint

Yes, matches KEGG results. Differences could occur because of pathway definitions/statistical methods from Reactome and Kegg. Some genes may also be mapped differently across databses.

##Section 5: GO Analysis

>Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Metabolic processes (p = 1.88E-134)
GO doesn't match exactly, but are consistent with KEGG. Go categories are broader than KEGG. Different statistical methods, pathways.
