---
title: "Class 10: Halloween mini project"
author: "Joshua Martin (PID: A18545389)"
format: pdf
toc: true
---

As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our data comes from the 538 website and is available as a CSV file:

## Data Import
```{r}
candy <- read.csv("candy-data.csv")
candy
```

```{r}
library(flextable)
flextable::flextable(head(candy, 10))
```


> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
```{r}
candy |>
  nrow()
```
```{r}
library(tidyverse)
candy %>%
  nrow()
```
```{r}
candy <- read.csv("candy-data.csv", row.names =1)
head(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
candy |>
  nrow()
```
```{r}
library(tidyverse)
candy %>%
  nrow()
```
```{r}
candy <- read.csv("candy-data.csv", row.names =1)
head(candy)
```


## 2. What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy in the dataset is Skittles
```{r}
candy["Skittles original", ]$winpercent
```

```{r}
library(dplyr)
candy |>
  filter(rownames(candy)=="Twix") |>
  select(winpercent)
```
```{r}
candy |>
  filter(rownames(candy)=="Almond Joy") |>
  select(winpercent)
```
## Quick overview of the dataset

```{r}
skimr::skim(candy)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Twix", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
skim(candy)
```
The variable winpercent stands out because the rest of mostly binary 0s and 1s. Winpercent represents a popularity score and is not categorical and thus is on a different numerical scale than the rest. The winpercent is on 0-100 scale the rest are 0-1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

That the data does not contain chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20)
```


> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```
```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
# 2. Find their winpercent values
# 3. Calculate the mean of these values

# 4-6. Do the same for fruit candy
# 7. Compare mean winpercents of chocolate vs fruity
# 8. Pick the highest as the winner

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)
choc.mean
```

```{r}
mean(candy[candy$chocolate==1,]$winpercent)
```

```{r}
fruity.inds <- candy$fruit==1
fruity.win <- candy[fruity.inds, ]$winpercent
fruity.mean <- mean(fruity.win)
fruity.mean
```

```{r}
mean(candy[candy$fruity==1,]$winpercent)
```

```{r}
candy |>
  filter(chocolate==1) |>
  select(winpercent)
```

```{r}
candy |>
  filter(fruity==1) |>
  select(winpercent)
```


> Q12. Is this differece statistically significant?

```{r}
t.test(choc.win, fruity.win)
```
## 3. Overall Candy Ranking

> Q13. What are the five least liked candy types in this set?

Nik L Nip
Boston Baked Beans
Chiclets
Super Bubble
Jawbusters
```{r}
head(candy[order(candy$winpercent),], n=5)
```
```{r}
candy %>% arrange(winpercent) %>% head(5)
```
```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
x <- c(5,1,10,4)
#sort(x)
order(x)
#(candy$winpercent)
```

```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```



> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(winpercent) |>
  tail(5)
```
```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Nik L Nip

## 4. Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent 

```{r}
my_cols[as.logical(candy$fruit)] <- "red"

ggplot(candy) +
  aes(x=winpercent,
      y=pricepercent,
      label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text()
```

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x=winpercent,
      y=pricepercent,
      label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 7)
```

Now that we've explored the dataset a little, we'll see how the variables interact with one another. We will plot a correlation matrix.

```{r}
library(corrplot)

cij <- cor(candy)
cij
corrplot(cij)
```
## 5. Principal Component Analysis

The function to use i called  `prcomp()` with an optional `scale=T/F` argument.
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Our main PCA result figure
```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)
```

We should also examine the variable "loadings"
or contrubutions of the orginional variables to the new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1 , rownames(pca$rotation)) +
  geom_col()
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])

```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Interactive plots that can be zoomes on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF :-(

```{r}
library(plotly)

#plotly(p)
```


